Methods that remove batch effects while retaining group differences may lead to exaggerated confidence in downstream analyses

### 2014-11-19 tex/manuscript.tex
Oppdatert manus (og diff)
Figur 1 og figurtekst må oppdateres, ellers tror jeg det meste skulle
være på plass nå.

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, as batch adjustments may otherwise bias, usually deflate, group differences when study groups are not evenly balanced across batches. Using a two-way ANOVA model to simultaneously estimate both group an batch effects is a natural approach in such cases. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. \COMMENT{The scientific community seems to be largely unaware of how this approach may lead to false discoveries.}{Har moderert påstanden noe, jfr. reviewer~2, pt.~6.}
}
\end{abstract}

@@

The detailed calculations for general linear models are presented in the appendix. Here, we summarise the results for two-way ANOVA based batch adjustments based on \eqref{eq:model} when there are $m$ batches, $M$ study groups, and $n_{ij}$ samples in batch $i$ and group $j$.

A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{q,n-q-1}$ distribution. However, if the data have been batch adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
\begin{equation}
F\sim\frac{\sum_{i=1}^q\lambda_i\chi^2_{1}}{\chi^2_{n-q-r-1}}
\approx\frac{\tilde q\tilde\sigma^2}{q\sigma^2}\cdot\left(1+\frac{r}{n-q-r-1}\right)\cdot\text{F}_{\tilde q,n-q-r-1}
\end{equation}
where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge q\sigma^2$ and $\tilde q\le q$ with equalities if and only if the groups are evenly represented in all batches.

The two factors both cause a bias in the $F$-statistic. The first factor, $\tilde q\tilde\sigma^2/q\sigma^2$, captures the unbalancedness of the design but is otherwise independent of the sample size. The second factor, $q-r/(n-q-r-1)$ captures loss of variance due to the centering of the data, but is primarily a problem when there are few samples in each batch. Finally, the degrees of freedom, $\tilde q$ and $n-q-r-1$, of the F-distribution are smaller than assumed by the $\text{F}_{q,n-q-1}$ distribution, and so the variance is larger.

As the sample size increases, assuming the distribution between groups and batches remain unchanged, the effect of the second factor and the second degree of freedom will diminish. However, the factor $\tilde q\tilde\sigma^2/q\sigma^2$ will remain, as will the reduced degree of freedom $\tilde q$.

@@

Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made (\citealp{Sandve2013}). In cases where data preparation is performed prior to, and separate from, the data analysis, as is now often the case, this is of particular importance, as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses. When assessing published findings, we frequently found it hard to determine how batch adjustments had been done, and occasionally suspected that study group had been used as covariate without this being stated. Science not only depends on determining sound methods and finding reliable results, but also on communicating sufficient detail to allow readers to assess the extent to which findings can be trusted. Failing to do so may lead cause methodological problems to pass undetected, but also to cast doubt on sound results. An increased focus on making research reproducible is therefore of great importance.

>>>>

\begin{abstract}
{Removal of, or adjustment for, batch effects or centre differences is generally required when such effects are present in data. In particular, when preparing microarray gene expression data from multiple cohorts, array platforms, or batches for later analyses, batch effects can have confounding effects. Many methods and tools exist for this purpose. One method, ComBat, which is part of the R package sva, is particularly popular due to its ability to remove batch differences even when batches are small and heterogeneous. It also has the option of preserving the difference between study groups, as batch adjustments may otherwise bias, usually deflate, group differences when study groups are not evenly balanced across batches. Using a two-way ANOVA model to simultaneously estimate both group an batch effects is a natural approach in such cases. Unfortunately, this frequently used and recommended approach may systematically induce incorrect group differences in downstream analyses when groups are distributed between the batches in an unbalanced manner. The scientific community seems to be largely unaware of how this approach may lead to false discoveries.
}
microarrays, data normalisation, batch effect, ANOVA, linear models
\end{abstract}

@@

The detailed calculations for general linear models are presented in the appendix. Here, we summarise the results for two-way ANOVA based batch adjustments based on \eqref{eq:model} when there are $m$ batches, $M$ study groups, and $n$ samples with $n_{ij}$ samples in batch $i$ and group $j$. Assuming the two-way ANOVA model for batch and group effects is non-degenerate, the degrees of freedom in the ANOVA model as described in the appendix will be $p=1$ for the intercept, $q=M-1$ for the groups, and $r=m-1$ for the batches.

A one-way ANOVA checking for group differences will assume that the $F$-statistic follows an $\text{F}_{M-1,n-M}$ distribution as group differences have $M-1$ degrees of freedom, while there $n-M$ degrees of freedom used to estimate the variance within groups.

However, if the data have been batches adjusted prior to this analysis using two-way ANOVA, the approximate distribution of this $F$-statistic, with correct first two momenta, will be
\begin{equation}
F\sim\frac{\sum_{i=1}^{M-1}\lambda_i\chi^2_{1}}{\chi^2_{n-M-m+1}}
\approx\frac{\tilde q\tilde\sigma^2}{(M-1)\sigma^2}\cdot\left(1+\frac{m-1}{n-M-m+1}\right)\cdot\text{F}_{\tilde q,n-M-m+1}
\end{equation}
where the eigenvalues $\lambda_i$, $\tilde q$, and $\tilde\sigma^2$ are as computed in the appendix and satisfy $\tilde q\tilde\sigma^2\ge (M-1)\sigma^2$ and $\tilde q\le M-1$ with equalities if and only if the groups are evenly represented in all batches.

The two factors both cause a bias in the $F$-statistic. The first factor, $\tilde q\tilde\sigma^2/(M-1)\sigma^2$, captures the unbalancedness of the design, but is otherwise independent of the sample size. The second factor, $1-(m-1)/(n-M-m+1)$, is due to loss of variance from centering of the data, which can be influential when there are few samples in each batch. Finally, the degrees of freedom, $\tilde q$ and $n-M-m+1$, of the F-distribution are smaller than assumed by the $\text{F}_{M-1,n-M}$ distribution, and so the variance is larger.

As the sample size increases, assuming the distribution between groups and batches remain unchanged, the effect of the second factor and the second degree of freedom will diminish. However, the factor $\tilde q\tilde\sigma^2/(M-1)\sigma^2$ will remain, as will the reduced degree of freedom $\tilde q$.

@@

Finally, we would like to emphasise the importance of proper description of how data has been prepared for analysis, and what corrections and adjustments have been made (\citealp{Sandve2013}). In cases where data preparation is performed prior to, and separate from, the data analysis, as is now often the case, this is of particular importance, as artifacts may be introduced in the data preparation which could influence the reliability of downstream analyses. When assessing published findings, we frequently found it hard to determine how batch adjustments had been done, and occasionally suspected that study group had been used as covariate without this being stated. Science not only depends on determining sound methods and finding reliable results, but also on communicating sufficient detail to allow readers to assess the extent to which findings can be trusted. Failing to do so may cause methodological problems to pass undetected, but also to cast doubt on sound results. An increased focus on making research reproducible is therefore of great importance.
